Modular analysis of gene expression data with R

### 2009-12-09 isanote/modular.tex
Added Sven's changes to the ISA application note.

\begin{abstract}
\section{Summary:}
Large sets of data, like expression profile from many samples, require
analytic tools to reduce their complexity. Classical (bi-)clustering
algorithms typically attribute elements (genes, arrays) to disjoint groups
(``clusters''). Yet, in some cases overlapping cluster assignments would suit
the biological reality much better.

The Iterative Signature Algorithm (ISA) was designed to overcome this and
other limitations of standard clustering algorithms. It aims to reduce the
complexity of very large sets of data by decomposing it into so-called
``modules''. In the context of gene expression data these modules consist of
subsets of genes that exhibit a coherent expression profile only over a
subset of microarray experiments. Genes and arrays may be attributed to
multiple modules and the level of required coherence can be varied resulting
in different ``resolutions'' of the modular mapping. Since the ISA does not
rely on the computation of correlation matrices (like many other tools), it
is extremely fast even for very large datsets.

In this short note, we introduce two BioConductor, \cite{BioC}, software
packages written in GNU~R: The~\Rpackage{isa2} includes a highly optimized
implementation of the ISA and~\Rpackage{eisa} provides a convenient interface
to run the ISA and visualize its output.

@@

% How the ISA works
The ISA can be applied to identify coherent substructures (i.e. modules)
from any rectangular matrix of data. To be specific we consider here the
case of transcriptomics data corresponding to a selection of gene-expression
profiles from a collection of samples. The method has been described in
detail in Refs~\cite{isamod,isa}. Here we only give a brief summary.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{isa2workflow3}
\caption{Work flow of a typical modular analysis study, with the
  \Rpackage{eisa} package (left); and four bicluster visualization
  techniques (right). The input of the \Rpackage{eisa} pipeline is a
  BioConductor \Rclass{ExpressionSet} object. This is normalized to
  $Z$-scores (1); the ISA iterations are initiated using ``smart'' or
  random seeds (2); and the biclusters are then merged and filtered
  based on the robustness measure (3). The \Rpackage{eisa} package
  contains tools to perform many GO, KEGG, etc. enrichment calculations
  quickly (4), and several bicluster visualization tools (5).
  Subfigures \subfig{A}, \subfig{B}, \subfig{C} and \subfig{D} were
generated
  using the acute lymphoblastic leukemia data set, see
  \cite{chiaretti04} and the \Rpackage{ALL} R package.
  \subfig{(A)} A condition plot for a simple bicluster. Each sample is
  represented as a bar, its height is its score in the bicluster.
  Samples above the red (none in this case) and samples
  below the green line are part of the biclusters. The coloring
  indicates cases and controls.
  \subfig{(B)} A module tree. Each rectangle is a bicluster; see the
  definition of the edges in the text. The modules are colored
  according to their Gene Ontology enrichment $p$-values, the codes of
  the enriched GO categories are shown in the top-left corner of the
  rectangles. The top-right corner shows the number of genes and
  conditions in the bicluster. The gene thresholds used for finding
  the modules is shown on the horizontal axes.
  \subfig{(C)} Heatmap for a single module, showing correlated
  expression for the genes and samples. The red lines are the gene and
  sample scores.
  \subfig{(D)} Profile plot, for a single module. On the top plot the
  expression of all samples is plotted, the samples of the module
  with black. The red and green lines show the mean of the two
  groups. The bottom plot is the same for the genes of the module.
}
\label{fig:workflow}
\end{figure*}

@@

% Batch correction
\emph{Batch correction.}
To study the global organization of a transcription program including
many aspects of transcriptional regulation one often combines several
microarray experiments into a single dataset. In such a case it is important
to properly normalize such data and reduce the bias due to constituent
datasets. Several methods address this challenge,
see e.g.~\cite{johnson07} for an algorithm that has a GNU R implementation.
% (S: Do we have a standard flag to call their software?)

@@

% The ISA iteration.
\emph{The ISA iteration.}
This step is essentially performing the ISA and finding the biclusters
from the starting seeds.

% Merging the modules.
\emph{Merging the modules.}
It is possible that more seeds converge to the same, or very similar
biclusters. This step eliminates such duplicates.

@@

The ISA and accompanying visualization tools are implemented in two R
packages. The \Rpackage{isa2} package contains the implementation of
the basic ISA itself; this package can be used to analyze any tabular
data. The \Rpackage{eisa} package builds on \Rpackage{isa2}. It adds
support to standard BioConductor gene expression data structures; and
contains gene expression specific visualization tools, see some of
these in Fig.~\ref{fig:workflow}.

Both the \Rpackage{isa2} and \Rpackage{eisa} packages support two
workflows. The \emph{simple workflow} involves a single R function
call and runs all ISA steps (1-3 on Fig.~\ref{fig:workflow}.) with
their default parameters.

@@

The \Rpackage{eisa} package implements a set of visualization
techniques for biclusters, some of these are specific to ISA, others
are not. Please see Fig.~\ref{fig:workflow} for some of these.

The \Rpackage{ExpressionView} R package, discussed in detail in
another application note, implements an interactive tool
for visualizing ISA (or other) overlapping biclusters.

@@

The \Rpackage{biclust} package, \cite{biclust}, implements a number of
biclustering algorithms in a unified framework. The \Rpackage{eisa}
package include tools to convert between \Rpackage{biclust} and ISA
biclusters. This allows the cross-talk of the functions in the two
packages.

@@

\paragraph{Funding\textcolon} The authors are greatful to the Swiss
Institute of Bioinformatics for support.
\paragraph{Conflict of interest\textcolon} None declared.

>>>>

\begin{abstract}
\section{Summary:}
Large sets of data, like expression profiles from many samples, require
analytic tools to reduce their complexity. Classical (bi-)clustering
algorithms typically attribute elements (genes, arrays) to disjoint groups
(``clusters''). Yet, in some cases overlapping cluster assignments would suit
the biological reality much better.

The Iterative Signature Algorithm (ISA) was designed to overcome this and
other limitations of standard clustering algorithms. It aims to reduce the
complexity of a very large set of data by decomposing it into so-called
``modules''. In the context of gene expression data these modules consist of
subsets of genes that exhibit a coherent expression profile only over a
subset of microarray experiments. Genes and arrays may be attributed to
multiple modules and the level of required coherence can be varied resulting
in different ``resolutions'' of the modular mapping. Since the ISA does not
rely on the computation of correlation matrices (like many other tools), it
is extremely fast even for very large datasets.

In this short note, we introduce two BioConductor \citep{BioC}, software
packages written in GNU~R: The~\Rpackage{isa2} includes a highly optimized
implementation of the ISA and~\Rpackage{eisa} provides a convenient interface
to run the ISA and visualize its output.

@@

% How the ISA works
The ISA can be applied to identify coherent substructures (i.e. modules)
from any rectangular matrix of data. To be specific we consider here the
case of transcriptomics data corresponding to a set of gene-expression
profiles from a collection of samples. The method has been described in
detail in Refs~\citep{isamod,isa}. Here we only give a brief summary.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{isa2workflow3}
\caption{Work flow of a typical modular analysis study, with the
  \Rpackage{eisa} package (left); and four bicluster visualization
  techniques (right). The input of the \Rpackage{eisa} pipeline is a
  BioConductor \Rclass{ExpressionSet} object. This is normalized to
  $Z$-scores (1); the ISA iterations are initiated using ``smart'' or
  random seeds (2); and the biclusters are then merged and filtered
  based on the robustness measure (3). The \Rpackage{eisa} package
  contains tools to perform many GO, KEGG, etc. enrichment calculations
  quickly (4), and several bicluster visualization tools (5).
  Subfigures \subfig{A}, \subfig{B}, \subfig{C} and \subfig{D} were
generated
  using the acute lymphoblastic leukemia data set, see
  \citep{chiaretti04} and the \Rpackage{ALL} R package.
  \subfig{(A)} A condition plot for a simple bicluster. Each sample is
  represented as a bar, its height is its score in the bicluster.
  Samples above the red (none in this case) and samples
  below the green line are part of the biclusters. The coloring
  indicates cases and controls.
  \subfig{(B)} A module tree. Each rectangle is a bicluster; see the
  definition of the edges in the text. The modules are colored
  according to their Gene Ontology enrichment $p$-values, the codes of
  the enriched GO categories are shown in the top-left corner of the
  rectangles. The top-right corner shows the number of genes and
  conditions in the bicluster. The gene thresholds used for finding
  the modules is shown on the horizontal axes.
  \subfig{(C)} Heatmap for a single module, showing correlated
  expression for the genes and samples. The red lines are the gene and
  sample scores.
  \subfig{(D)} Profile plot, for a single module. On the top plot the
  expression of all samples is plotted, the samples of the module
  with black. The red and green lines show the mean of the two
  groups. The bottom plot is the same for the genes of the module.
}
\label{fig:workflow}
\end{figure*}

@@

% Batch correction
\emph{Batch correction.}
To study the global organization of a transcription program including
many aspects of transcriptional regulation one often combines several
microarray experiments into a single dataset. In such a case it is important
to properly normalize such data and reduce the bias due to constituent
datasets. Several methods address this challenge,
see e.g.~\citep{johnson07} for an algorithm that has a GNU R
implementation. 
% (S: Do we have a standard flag to call their software?)

@@

% The ISA iteration.
\emph{The ISA iteration.}
This step is the core of the ISA and finds biclusters from the
starting seeds.

% Merging the modules.
\emph{Merging the modules.}
It is possible that several seeds converge to the same, or very similar
biclusters. This step eliminates such duplicates.

@@

The ISA and accompanying visualization tools are implemented in two R
packages. The \Rpackage{isa2} package contains the implementation of
the basic ISA itself; this package can be used to analyze any tabular
data. The \Rpackage{eisa} package builds on \Rpackage{isa2}. It adds
support to standard BioConductor data structures; and
contains gene expression specific visualization tools (see 
Fig.~\ref{fig:workflow} for examples).

Both the \Rpackage{isa2} and \Rpackage{eisa} packages support two
workflows. The \emph{simple workflow} involves a single R function
call and runs all ISA steps (1-3 in Fig.~\ref{fig:workflow}.) with
their default parameters.

@@

The \Rpackage{eisa} package implements a set of visualization
techniques for biclusters, one of which is specific to ISA (see
Fig.~\ref{fig:workflow} for examples).

% The \Rpackage{ExpressionView} R package, discussed in detail in
% another application note, implements an interactive tool
% for visualizing ISA (or other) overlapping biclusters.

@@

The \Rpackage{biclust} package, \citep{biclust}, implements a number of
biclustering algorithms in a unified framework. The \Rpackage{eisa}
package include tools to convert between \Rpackage{biclust} and ISA
biclusters. This allows the cross-talk of the functions in the two
packages.

@@

\paragraph{Funding\textcolon} The authors are greatful to the Swiss
Institute of Bioinformatics, the Swiss National Science Foundation
(3100AO-116323/1), and the European Framework Project 6 (through
the EuroDia and AnEuploidy projects).
\paragraph{Conflict of interest\textcolon} None declared.
